---
title: "hw3_rw3031"
author: "Polly Wu (rw3031)"
date: "2024-10-05"
output: github_document
---

## library and visual theme set up

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal()+theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_color_viridis_d
```

# Problem 1 

load the NY NOAA dataset
```{r}
library(p8105.datasets)
data("ny_noaa")
```

### explore the NY NOAA dataset

```{r}
nrow(ny_noaa)
ncol(ny_noaa)
summary(ny_noaa)

ny_noaa |> 
  ggplot(aes(x = prcp)) + 
  geom_histogram()

ny_noaa |> 
  ggplot(aes(x = snow)) + 
  geom_histogram()

ny_noaa |> 
  ggplot(aes(x = snwd)) + 
  geom_histogram()
```
There are 2595176 observations and 7 variables (id, date, prcp, snow, snwd, 
tmax, tmin) in dataset. The variable tmax and tmin are imported as character variable.

The earliest date for the data is 1981-01-01 and the most recent date is 2012-12-31.

Precipitation ranges from 0 to 22860 (tenths of mm) and 145838 observations do not 
report precipitation data. Snowfall range from -13 mm to 10160 mm, and 381221 
observations' data on snowfall is missing, Snow depth range from 0 to 9195 mmm 
and 591786 observations do not have snow depth data. All three variables are highly
right skewed since there are a lot of days that there is no rain or snow

```{r}
ny_noaa = 
  ny_noaa|>
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))

summary(ny_noaa)
```

I change the character variable tmax and tmin into numeric instead. 

There are more than 1100,000 observations that have the maximum and minimum temperature 
missing within the dataset which is more than 40% of the observations. 

### cleaning the NY_NOAA data 

Create separate variables for year, month, and day. 
```{r}
ny_noaa=
ny_noaa|>
  mutate(year = year(date),
         month = month(date),
         day = day(date))
```

```{r}
ny_noaa |> 
  ggplot(aes(x = snow)) + 
  geom_histogram()
```
For snowfall the most commonly observed value is 0 since except for cold winter time
there is not any snow. 

### average temperatue in January and July

```{r}
ny_noaa|>
  filter(month ==1 | month == 7)|>
  group_by(year,month)|>
  summarize(avg_tmax = mean(tmax, na.rm = TRUE))|>
  ggplot(aes(x = year, y=avg_tmax))+
  geom_point()+geom_line()+
  facet_grid(.~month)
```

The average maximum temperature in January and July fluctuates a lot. The years with higher tmax in January tend to have higher tmax in July as well. 

### tmax and tmin and distribution of snowfall

```{r}
tmax_tmin_p=
ny_noaa|>
  ggplot(aes(x= tmax, y=tmin))+
  geom_bin2d()+
  theme(legend.position = "right")

snow_p=
ny_noaa|>
  filter(snow>0, snow<100)|>
  group_by(year)|>
  ggplot(aes(x=snow))+
  geom_histogram()

tmax_tmin_p + snow_p
```

# Problem 2 

import the datasets 

```{r}
demo =
  read_csv(file = "./nhanes_covar.csv", na = c(".", "NA", ""))|>
  janitor::clean_names()


accel =
  read_csv(file = "./nhanes_accel.csv", na = c(".", "NA", ""))|>
  janitor::clean_names()|>
  pivot_longer(
    min1
  )
```



# Problem 3 
